---
---
---

```{r setup}
#install.packages("dplyr")
library(dplyr)
library(ggplot2)
```

We will be practicing with `dplyr` functions using the `starwars` dataset:

```{r}
head(starwars)
starwars
?starwars
```

## Subset rows with `filter`

`filter()` allows use to subset the **rows** of a dataset based on the values

```{r}
## Filter for rows with human Star Wars characters
filter(starwars, species == "Human")

## save to a new R object
humans <- filter(starwars, species == "Human")
humans
filter(humans, sex == "male")

## filter multiple columns at once
filter(starwars, species == "Human", sex == "male")
```

We can use different comparison operators to filter:

-   `!` is "not"

-   `==` is "equal"

-   `&` is "and"

-   `|` is "or"

-   `<` is "less than"

-   `>` is "greater than"

-   `is.na()` returns TRUE is value is NA

```{r}
## return characters with a missing mass
filter(starwars, is.na(mass))
```

```{r}
## How to get all characters that are NOT human?
filter(starwars, species != "Human")
```

variable_name `%in%` vector_of_values returns TRUE if the value of variable_name is equal to one of the values in vector_values. This is a shortcut for variable_name == vector_of_values[1] or variable_name == vector_of_values[2], etc.

```{r}
skin_color <- "gray"
skin_color %in% c("yellow", "fair", "tan") #returns FALSE because "gray" is not in vector of colors

## How to filter for characters with brown or blond hair?
filter(starwars, hair_color %in% c("blond","brown"))
```

grepl("characters_of_interest", variable_name) returns TRUE if the value of variable_name contains characters_of_interest.

```{r}
filter(starwars, grepl("blond", hair_color))

## How to get all characters with blue in their eyes?
filter(starwars, grepl("blue", eye_color))
```

You can use `<`, `>`, `<=`, `>=` to compare numeric variables or categorical variables

-   nominal categorical variables (character vectors) are ranked alphabetically

-   ordinal categorical variables (factors) are ranked based on the built-in order

```{r}
## starwars characters with a name starting with letters before "D"
filter(starwars, name < "D") #ranking is based alphabetically

"D" < "d" #FALSE
"d" < "D" #TRUE
```

between(x, left, right) is a shorthand for x \>= left & x \<= right

```{r}
between(5,1,7) #TRUE because 5 is in between 1 and 7
filter(starwars, between(mass, 50, 100))
```

### Practice

Find all starwars characters that

1.  Are missing a hair color
2.  Are shorter than 100cm and are not droids
3.  Were born between years 100 and 200
4.  Another of choice

```{r}
filter(starwars, is.na(hair_color))

filter(starwars, height < 100, species != "Droid")

filter(starwars, between(birth_year, 100, 200))

filter(starwars, mass > 100, height > 150, sex == "male")
```

## Order rows with `arrange`

`arrange()` reorders rows based on the values of column(s)

```{r}
## order by character names
arrange(starwars, name)
# order of character values: aAbBcCdD

## order by species and birth year
arrange(starwars, species, birth_year)

## use desc() to get the birth years in descending order
arrange(starwars, species, desc(birth_year))
```

### Practice

1.  Arrange from tallest to shortest characters
2.  Arrange from shortest to tallest by species
3.  Alphabetize the characters by their hair color then eye color

```{r}
arrange(starwars, desc(height))

arrange(starwars, species, height)

arrange(starwars, hair_color, eye_color)
```

## Subset columns with `select()`

```{r}
select(starwars, hair_color, skin_color, eye_color)
character_colors <- select(starwars, hair_color, skin_color, eye_color)
character_colors
```

We can also select multiple columns in a row

```{r}
# all columns between name and eye_color (inclusive)
select(starwars, name:eye_color)
```

We can also remove columns by name

```{r}
select(starwars, -films, -vehicles, -starships)
select(starwars, -c(films, vehicles, starships))
select(starwars, -c(films:starships))
```

### Helpful Functions

`everything()` is a shortcut for "everything else"

```{r}
select(starwars, species, everything()) # put species first, then all other columns

## alphabetize columns
starwars_names <- sort(names(starwars))
select(starwars, starwars_names)
```

`starts_with("abc")` matches column names that start with "abc"

`ends_with("abc")` matches column names that end with "abc"

`contains("abc")` matches column names that contain "abc"

```{r}
select(starwars, starts_with("h"))

select(starwars, contains("or"))
```

### Practice

1.  Select each character's homeworld and species. What question might this subset of variables help answer?
2.  Brainstorm as many ways as possible to select `hair_color`, `skin_color`, `eye_color`, and `birth_year`

```{r}
select(starwars, homeworld, species)
# a question this subset might answer is: What is the percentage of each known species of characters from Star Wars on each planet?

select(starwars, hair_color, skin_color, eye_color, birth_year)
select(starwars, c(hair_color:birth_year))
select(starwars, contains("_"))
select(starwars, contains("color"), birth_year)
```

## Add new variables with `mutate`

```{r}
starwars_small <- select(starwars, height, mass, species)
starwars_small

mutate(starwars_small,
       height_m = height/100)
```

Can also create multiple variables at once and we can refer to variables we just created.

```{r}
mutate(starwars_small,
       height_m = height/100,
       bmi = mass/height_m^2)

## only keep new variables
transmute(starwars_small,
       height_m = height/100,
       bmi = mass/height_m^2)
```

Summary statistics give one value for the entire column: `sum(), mean()`

```{r}
mutate(starwars_small,
       sum = sum(mass, na.rm=T))

mutate(starwars_small,
       row_addition = height + mass,
       summary_sum = sum(height, mass, na.rm=T))
```

Group by a variable using the `.by` parameter

```{r}
mutate(starwars_small,
       species_mass = mean(mass,na.rm=T),
       .by = species)
```

Change the type of variable

```{r}
mutate(starwars_small,
       mass = as.character(mass))
```

### Practice

1.  Convert mass to pounds (1kg is 2.2lbs)

```{r}
mutate(starwars_small,
       mass_lbs = mass * 2.2)
```

## Summarize data with `summarize`

```{r}
summarize(starwars,
          mean_mass = mean(mass, na.rm=T))

# group by species
summarize(starwars,
          mean_mass = mean(mass, na.rm=T),
          .by = species)
```

The only columns kept are the columns you group by and the columns you create

`n()` is a helpful function that returns the number of rows in a group

```{r}
summarize(starwars,
          mean_mass = mean(mass, na.rm=T),
          count = n(),
          .by = c(species,sex))
```
